% =====================================================================
% Conclusion --- preliminary empirical fill based on the v3 logged KNMI run.
% Simulation and bootstrap inference are intentionally marked as TBD.
% =====================================================================
\section{Conclusion}\label{sec:conclusion}

This thesis asked whether the spatial range of extremal dependence in Dutch daily maximum wind gusts differs between winter and summer. To answer it, a season-specific Brown--Resnick max-stable process was fitted to \(N=33\) KNMI stations by pairwise composite likelihood, separately for winter (DJF) and summer (JJA), after standardising each station--season margin to the unit-Fr\'echet scale. The seasonal comparison was framed through the difference in dependence ranges \(\Delta_\rho=\rho_{\mathrm W}-\rho_{\mathrm S}\), with the implied extremal-coefficient curves \(\theta_c(d)\) placing that difference on the dependence-versus-distance scale. Uncertainty is intended to be quantified by a season-year block bootstrap, which preserves the within-storm serial dependence of consecutive daily maxima, and the estimator and the one-sided test are intended to be validated by a Monte Carlo study under the observed station geometry.

The simulation study is still to be determined. In the final version, it should establish the finite-sample behaviour of the pairwise composite-likelihood estimator under a known Brown--Resnick data-generating process. In particular, it should report whether the estimator recovers the seasonal ranges and smoothnesses with acceptable bias and RMSE, whether \(\widehat\Delta_\rho\) is centred near its true value under both the null and alternative designs, and whether the season-year block-bootstrap test has approximately nominal size and useful power. Until those experiments have been run, this thesis does not make a numerical simulation claim.

For the empirical KNMI application, the preliminary non-bootstrap fit gives \(\widehat\rho_{\mathrm W}=418.76\) km and \(\widehat\rho_{\mathrm S}=76.98\) km, so that \(\widehat\Delta_\rho=341.79\) km. The corresponding smoothness estimates are almost identical, with \(\widehat\alpha_{\mathrm W}=0.594\) and \(\widehat\alpha_{\mathrm S}=0.597\). The point estimates therefore indicate that the seasonal difference is mainly a difference in the spatial range of extremal dependence rather than in the shape of the distance-decay curve. Substantively, this supports the central hypothesis of the thesis: Dutch winter wind-gust extremes appear to have a much wider spatial footprint than summer wind-gust extremes, consistent with the contrast between large-scale winter storm systems and more localised summer convective gusts.

This empirical conclusion remains preliminary for two reasons. First, the current estimates come from a non-bootstrap run with the fast starting grid, so the final version should verify the point estimates using the full starting grid. Second, the formal one-sided test of \(H_0:\Delta_\rho=0\) has not yet been completed, because the season-year block-bootstrap interval and evidence measure are still to be determined. Therefore, the current conclusion should be read as a strong descriptive point-estimate result, not yet as a final inferential rejection statement.

Several limitations qualify the analysis. The Brown--Resnick model is asymptotically dependent at all finite distances, so it can overstate dependence at the highest levels if the gusts are in fact asymptotically independent; the nonparametric diagnostics are included precisely to detect this, and the conclusion is conditional on their outcome. Winter and summer are treated as approximately independent because they are estimated from disjoint days, though low-frequency interannual climate variation could induce some residual dependence. The margins are standardised by fitting a GEV to each station--season sample of daily maxima, which is a transformation device rather than a claim that daily maxima are exactly GEV-distributed, and an isotropic variogram is assumed throughout. Finally, the bootstrap-performance study is still pending, so the finite-sample calibration of the final test remains to be established.

These limitations point to natural extensions. Anisotropic and space-time Brown--Resnick models, such as the anisotropic space-time specification of Buhl and Kl\"uppelberg (2016), would relax the isotropy and single-snapshot assumptions and let the footprint depend on direction and on storm motion. A single covariate-dependent max-stable model, with season entering as a covariate rather than through separate fits, would test the seasonal contrast within one model and could incorporate further covariates such as storm type or large-scale circulation indices. Where the diagnostics indicate asymptotic independence, sub-asymptotic and asymptotically independent models surveyed by Huser and Wadsworth (2020) would describe the joint tail more faithfully than a max-stable process. Threshold-based margins, a finer seasonal resolution, and larger Monte Carlo and bootstrap budgets are further refinements. Taken together, these would extend the season-by-season, isotropic design adopted here towards a fuller account of how the spatial extent of damaging wind gusts varies across the year in the Netherlands.
